Memory dynamics for naturalistic experiences

### 2018-07-16 main.tex
more work on intro

% \title{Capturing the geometric structure of our experiences and how we remember them}
\title{The fundamental shape of dynamic experiences is preserved in episodic memory}
\author{Andrew C. Heusser \& Jeremy R. Manning}
% \affiliation{Dartmouth College}

@@

A defining feature of naturalistic experiences is the temporal structure with which they unfold. Our experiences in the real world are typically autocorrelated in space and time on short timescales. In other words, the contents of an experience are often similar from moment to moment and change gradually. For example, many of us spend our mornings/evenings at home and much of our days at work (e.g. an office). While the specific happenings within these contexts differs, they are (usually) highly correlated and predictable over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The perceptual contexts before and after lunch are highly similar, but separated in time. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely important if we wish to understand how our memory systems remember our everyday experiences.

Despite the continuous and (often) gradually changing nature of naturalistic experiences, prior research suggests that our memories for those same experiences may be more descrete and centered around particular ``events'' \citep{Radv12, RadvZack11, BrunEtal18}. Short timescale autocorrelations in the contents of an experience may cause information to become associatively linked in memory (and thus treated as a single event), while sharp changes (i.e. ``event boundaries'') may segment continuous experiences into events \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  Furthermore, brain activity while viewing naturalistic stimuli displays a block-diagonal structure, and the neurally-derived segments are correlated with judgements of event boundaries \citep{BaldEtal17}.  However, whether people naturally recount their experiences in as events has not been explored.

% We propose that variability in the contents of an experience over time define a ``stimulus-space''. In this space, timepoints where the contents of experience are more similar are represented as points that are closer together. Timepoints with relatively more distinct content are represented further apart. The moment-by-moment flow of content through this space can be represented as a ``stimulus trajectory''.
%
% As a consequence of this structure, we hypothesized that verbally recounted memories for those experiences would mirror this shape.

To study the temporal and semantic aspects of memories for naturalistic experiences, we analyzed an open dataset in which participants view and then verbally recall an episode of the BBC series \textit{Sherlock}  \citep{ChenEtal17}. We Furthermore, we hypothesized that the stimulus To test these hypotheses, we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17} to capture the temporal structure present in naturalistic experiences. We then use this model to characterize the shape of the stimulus as well as verbally recalled memories.

>>>>

% \title{Capturing the geometric structure of our experiences and how we remember them}
% \title{The fundamental shape of naturalistic experiences is preserved in episodic memory}
\title{Memory dynamics for naturalistic experiences}
\author{Andrew C. Heusser \& Jeremy R. Manning}
% \affiliation{Dartmouth College}

@@

Human memory studies typically assess memory for the individual contents of an experience, such as the people, places and things/items encountered. However, real life experiences can also be described in terms of their temporal structure (i.e. how the contents relate over time).  Naturalistic experiences are typically autocorrelated in space and time on short timescales: the contents of an experience are often similar from moment to moment and change gradually. For example, many of us spend our mornings/evenings at home and much of our days at work (e.g. an office). While the specific happenings within these contexts differ, they are often highly correlated (and thus predictable) over time. Furthermore, our experiences can also be characterized by longer timescale correlations. For example, consider returning to your office after taking a lunch break. The spatial contexts before and after lunch are highly similar, but separated in time. Thus, while evaluating the ability to recall specific contents of an experience is one way to assess memory, another important dimension to consider is the temporal structure of the remembered experience, and how the temporal structure of the memory relates to the original experience. However, these recurrent and gradually changing temporal dynamics are not typically present in traditional memory studies, but are likely critical if we wish to understand how our memory systems remember our life experiences.

Despite the continuous and (often) gradually changing nature of naturalistic experiences, prior research suggests that our memories for those same experiences may be more descrete and organized around particular ``events'' \citep{Radv12, RadvZack11, BrunEtal18}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience (i.e. ``event boundaries'') may lead to discontinuous in memory \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people naturally recount their experiences in a way that is structured around events is not known.

In this study, we analyzed an open dataset in which participants view and then verbally recount an episode of the BBC series \textit{Sherlock} \citep{ChenEtal17}. To capture the temporal structure present in naturalistic experiences, we developed a novel computational approach based on topic modeling \citep{BleiEtal03} and hidden Markov models \citep{Rabi89, BaldEtal17}. We propose that the temporal structure of an experience gives it a unique geometric ``shape'' or ``trajectory'', defined by how the contents of the experience relate and change over time. We hypothesized that when recounting naturalistic experiences, the precise details that are recapitulated and language used may differ across people, but the trajectory of the stimulus is generally preserved.
